---
title: In-memory interoperability
engine: knitr
---

## Reticulate: basic functionality
Reticulate is a foreign function interface in R to Python.

### Interactive sessions
One of the most useful ways to take advantage of in-memory interoperability is to use it in interactive sessions, where you're exploring the data and want to try out some functions non-native to your language of choice.

Jupyter notebooks (and some other notebooks) make this possible from the Python side: using IPython line and cell magic and rpy2, you can easily run an R jupyter cell in your notebooks.

```{python show_magic, eval=FALSE}
%load_ext rpy2.ipython  # line magic that loads the rpy2 ipython extension.
                        # this extension allows the use of the following cell magic

%%R -i input -o output  # this line allows to specify inputs (which will be converted to R objects) and outputs (which will be converted back to Python objects) 
                        # this line is put at the start of a cell
                        # the rest of the cell will be able to be ran as R code

```

### Reticulate basics


## Usecase

### Usecase: theoretical reticulate example

```{r read_in}
library(anndata)

adata_path <- "../usecase/data/sc_counts_subset.h5ad"
adata <- anndata::read_h5ad(adata_path)
```

Subset to a single small molecule and control for computational efficiency:

```{r select_sm_celltype}
library(dplyr)

sm_name <- "Belinostat"
control_name <- "Dimethyl Sulfoxide"

# subset obs
adata <- adata[adata$obs$sm_name %in% c(control_name, sm_name), adata$var$highly_variable]
```

```{r import_pandas}
library(reticulate)
pd <- import("pandas", convert = FALSE)

counts <- as.matrix(adata$X)
```

Combine data in a single data frame and compute pseudobulk
This is a literal translation of the Python code. It is however absolute madness to construct a pandas dataframe in R instead of using just an R dataframe. We basically just needed a matrix with rownames and columnames.

We provide the rest of the code as a theoretical example, but please reflect if you want to try something similar.

We will however showcase a useful application of reticulate.

```{r compute_pseudobulk, eval = FALSE}

combined <- pd$DataFrame(
  counts,
  index = adata$obs["plate_well_celltype_reannotated"],
  columns = adata$var_names
)
cr <- py_to_r(combined)

# we lost the rownames
rownames(cr) <- adata$obs_names
cr["celltype"] <- adata$obs["plate_well_celltype_reannotated"]

pb_X <- group_by(cr, celltype) %>% summarise(across(where(is.numeric), sum))
```

```{r pb_obs_r, eval = FALSE}
pb_obs <- adata$obs[c("sm_name", "cell_type", "plate_name", "well", "plate_well_celltype_reannotated")]
pb_obs <- pb_obs[!duplicated(pb_obs), ]
```

```{python pb_obs_py, eval = FALSE}
pb_obs = adata.obs[["sm_name", "cell_type", "plate_name", "well"]].copy()
pb_obs.index = adata.obs["plate_well_celltype_reannotated"]
pb_obs = pb_obs.drop_duplicates()
```

```{r pb_anndata, eval = FALSE}
select_X <- pb_X[pb_X$celltype %in% pb_obs$plate_well_celltype_reannotated, ]
select_X <- select_X %>% select(-"celltype")

pb_adata <- anndata::AnnData(
  X = select_X,
  obs = pb_obs,
  var = adata$var
)
```

<!-- note: don't remove the `eval=FALSE` for this one, as not to overwrite the usecase data -->
```{r store_pseudobulk, eval=FALSE}
write_h5ad(pb_adata, "../usecase/data/pseudobulk.h5ad")

```

### Usecase: useful reticulate example
